DCMIP2016, Part 1: Models and Equation Sets

### 2017-04-15 DCMIP2016-Part1-Models.tex
- More revisions - Added OLAM text

\begin{abstract}
Atmospheric dynamical cores are a core component of global atmospheric modeling systems, and are responsible for capturing the dynamical behavior of the Earth's atmosphere via numerical integration of the Navi\'er-Stokes equations.  These systems have existed in one form or another for over half of a century, with the earliest strategies now evolved into a complex ecosystem of algorithms and computational strategies.  In essence, no two dynamical cores are alike, and their individual successes suggest that no perfect model exists.  To better understand modern dynamical cores, this paper aims to provide a comprehensive review of their design, including model grids, relevant equation sets, numerical stabilization techniques and idealized physics routines.  Examples are taken from operational dynamical cores that participated in the 2016 Dynamical Core Model Intercomparison Project (DCMIP) workshop and summer school.
\end{abstract}

@@

The Dynamical Core Model Intercomparison Project (DCMIP) is an ongoing effort targeting the intercomparison of a fundamental component of global atmospheric modeling systems: the dynamical core.  Although this component's role is simply to solve the equations of fluid motion (the Navi\'er-Stokes equations) governing atmospheric dynamics, there are numerous confounding factors that arise as a consequence of compromises that are required to make simulation computationally feasible.  These factors include the choice of model grid, choice of prognostic equations, vertical coordinates, representation of topography, numerical method, temporal discretization, physics/dynamics coupling frequency, and the manner in which artificial diffusion, filters and/or energy/mass fixers are applied.

To advance the intercomparison project and provide a unique educational opportunity for students, DCMIP hosted a multidisciplinary two-week summer school and model intercomparison project, held at the National Center for Atmospheric Research (NCAR) in June 2016, that invited graduate students, postdocs, atmospheric modelers, expert lecturers and computer specialists to create a stimulating, unique and hands-on driven learning environment. It was built on earlier DCMIP workshops (held in 2008 and 2012), and previous intercomparison efforts by addressing key outstanding issues in global atmospheric models, incorporate international participation, and provide a unique training experience for the future generation of climate scientists. Special attention was paid to the role of simplified physical parameterizations, physics-dynamics coupling, non-hydrostatic atmospheric modeling and variable-resolution global modeling. The summer school and model intercomparison project promoted active learning, innovation, discovery, mentorship and the integration of science and education.  Modeling groups were then invited to contribute model descriptions and results to the intercomparison effort for publication.

The summer school directly benefited its participants by providing a unique educational experience and an opportunity to interact with modeling teams from around the world.  The workshop is expected to have further repercussions on the development of operational atmospheric modeling systems, by giving modeling groups an opportunity to assess and intercompare their models with other advanced modeling systems.  Modeling groups have leveraged this information to improve their own models, which will in turn positively impact the quality of weather and climate simulations going forward.

The workshop has advanced our knowledge of (1) the relative behaviors exhibited by atmospheric dynamical cores, (2) differences that arise among mechanisms for coupling the physical parameterizations and dynamical core, and (3) the impacts of variable-resolution refinement regions and transition zones in global atmospheric simulations.  Notably, the use of idealized test cases isolating specific phenomena gave us a unique opportunity to assess specific differences that arise due to the choice of dynamical core.  A key outcome of the workshop was the development of a standard test case suite and benchmark set of simulations that can be used for assessment of any future dynamical core.  The test cases introduced in the 2016 workshop build on the previous DCMIP test case suites \citep{jablonowski2008idealized, ullrich2012dynamical} with tests that now incorporate simplified moist physics.

This paper is the first in a series of papers documenting the results of this workshop.  This paper is intended to provide a review of the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling and provide an introduction to the intercomparison strategy that has been employed by DCMIP.  For easy reference, a list of mathematical symbols that are employed in this paper (and subsequent DCMIP papers) is given in Table \ref{tab:symbols}.  Section \ref{sec:DynamicalCores} provides a brief overview of each of the participating models.  The middle of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  Section \ref{sec:ModelGrids} describes model grids that have been employed by modern modeling systems; section \ref{sec:EquationSets} describes different formulations of the moist dynamical equations; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; and section \ref{sec:TemporalDiscretizations} describes temporal discretizations.  Conclusions and discussion then follow in section \ref{sec:Conclusions}.

\begin{table}[h]
\caption{List of symbols used by DCMIP} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind \\
$v$ & Meridional wind \\
$w$ & Vertical velocity \\
$\vb{u}$ & 3D wind vector \\
$\vb{u}_h$ & Horizontal wind vector \\
$\vb{v}_h$ & Horizontal wind vector with covariant components \\
$\omega$ & Vertical pressure velocity  \\
$D$ & Divergence of the horizontal wind vector\\
$\zeta$ & Vertical component of relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$e$ & Internal energy \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$\rho_s$ & Pseudo-density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\theta$ & Potential temperature \\
$\theta_v$ & Virtual potential temperature \\
$q$ & Specific humidity \\
$P_{ls}$ & Large-scale precipitation rate \\
$q$ & Water vapor mixing ratio \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
$q_i$ & General tracer mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

This section provide a brief overview of key discretization choices, along with unique features or design specifications from participating dynamical cores.  Further details on these choices can be found in subsequent sections.  Simulation results and model descriptions have been submitted from ten dynamical cores (see Table \ref{tab:Models}).  Although all models solve the Euler equations (or a close approximation of these equations), essentially every model employs a different formulation with a distinct set of prognostic variables.  The derivation of many of these formulations can be found in Appendix \ref{sec:EquationSets}.  The prognostic variables employed and horizontal discretizations for these dynamical cores are summarized in Table \ref{tab:ModelsHorizontal}.  The vertical coordinate chosen by these models and vertical staggering of variables is summarized in Table \ref{tab:ModelsVertical}.  Standard options for diffusion, stabilization, filters, or fixers along with the temporal discretization for these models is summarized in Table \ref{tab:ModelsTemporalDiffusion}.

\begin{table}[h]
\caption{Participating modeling centers and associated dynamical cores that have submitted simulation results.} \label{tab:Models}
\begin{center}
\begin{tabular}{cll}
\hline Acronym & Model Name & Modeling Center or Group \\ \hline 
CSU & Colorado State University Model & Colorado State University, USA \\
DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
FV$^3$ & GFDL Finite-Volume Cubed-Sphere Dynamical Core & Geophysical Fluid Dynamics Laboratory, USA \\
FVM & Finite Volume Module of the Integrated Forecasting System & European Centre for Medium-Range Weather Forecasts \\
GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
HOMME & High-Order Method Modeling Environment & Sandia National Laboratories / University of Colorado, Boulder \\
ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie, Germany \\
MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research, USA \\
NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
Tempest & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis, USA \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[h]
\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.  Equation set indicates whether a model is hydrostatic (H) or nonhydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).  Only two numerical methods are represented among participating models, namely finite-volume (FV) and spectral-element (SE).  More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Equation set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
& & & & method & staggering \\ \hline
CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
DYNAMICO & H/NH {\color{red}(D)?} & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
FVM & NH (D) & $\rho_d$, $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d \theta$, ($\Phi$), $\rho_d q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
GEM & NH & $\vb{u}_h$, $w$, {\color{red}$\log \partial s / \partial \zeta$}, $\log T_v$, $q$ & Yin-yang (\S \ref{sec:grid_yinyang}) & FV & C-grid  \\
HOMME & H & $\vb{u}_h$, $\rho_s$, $T_v$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
ICON & NH & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[h]
\caption{Vertical coordinates and staggering for participating dynamical cores.  More details on vertical coordinates can be found in section \ref{sec:VerticalCoordinates}, and details on vertical staggerings can be found in section \ref{sec:VerticalStaggering}.} \label{tab:ModelsVertical}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Vertical Staggering & Vertical Coordinate \\ \hline 
CSU & Lorenz & fixed height \\
DYNAMICO & Lorenz & {\color{red}floating mass?} \\
FV$^3$ & Lorenz & floating mass \\
FVM & Co-located & (floating) height \\
GEM & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering}) & log pressure (\S \ref{sec:GEM_zeta})  \\
HOMME & Co-located & floating mass \\
ICON & Lorenz & fixed height \\
MPAS & Lorenz & fixed height \\
NICAM & Lorenz & fixed height \\
Tempest & Lorenz & fixed height \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[h]
\caption{Standard options for diffusion, stabilization, filters, or fixers in participating dynamical cores, detailed in section \ref{sec:DiffusionStabilization}, and temporal discretization, detailed in section \ref{sec:TemporalDiscretizations}.} \label{tab:ModelsTemporalDiffusion}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Standard options for diffusion, & Temporal discretization \\
& stabilization, filters, or fixers & \\ \hline 
CSU & 4th-order horizontal hyperdiffusion & 3rd-order Adams-Bashforth (AB3) \\
DYNAMICO & {\color{red}?} & ARK232 \citep{giraldo2013implicit} \\
FV$^3$ & Divergence damping, monotonic limiting, hyperviscosity & Forward-backwards \citep{LR1997QJR} \\
FVM & Momentum dissipation, scalar diffusion, absorbing layer & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
HOMME & Hyperdiffusion & KGU53 \citep{guerra2016high} \\
ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
MPAS & Smagorinsky, hyperdiffusion & Split-explicit \citep{klemp2007conservative} \\
NICAM & 3D divergence damping, Smagornisky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
Tempest & Hyperdiffusion & ARS232 \citep{ascher1997implicit} \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

@@

The GEM model \citep{Girard2014} is used for operational forecasting at Environment and Climate Change Canada.  The horizontal discretization uses the Yin-Yang grid \citep{kageama2004yinyang}, with Arakawa C-grid staggering.
The vertical coordinate is a unique hybrid terrain-following coordinate of a log-hydrostatic-pressure type
and the vertical discretization is based on the Charney-Phillips grid (see section \ref{sec:VerticalStaggering}).
A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.
It gives rise to an {\it{iterative}} process where each step requires the solution of a
linear system of equations that is reduced to a Helmholtz problem for one composite variable.
For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method \citep{Qaddouri2008schwarz}.
The dynamics and physics are time split.
To eliminate numerical noise, an explicit hyperviscosity is employed for wind components and tracers
via applications of the Laplacian operator.
Viscosity operations are applied after the completion of the physics time step.

@@

The ICON model \citep{zangl2015icon} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain-following height-based Lorenz grid.  Prognostic horizontal velocities are stored as normal wind components $v_n$ at the edge mid-points of full levels.  Prognostic vertical wind $w$ is stored at the circumcenters of the triangles on half levels. The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves. Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics. One complete time step typically includes 5 dynamical sub-steps. The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport. For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells. Further divergence damping is applied to the normal wind at every sub-step. Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves. The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinsky scheme with a background 4th order Laplacian diffusion operator. For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion. The numerical methods have been chosen for high numerical efficiency, and they rely on next-neighbor communication only, thus allowing massive parallelization.

@@

@@

@@

\section{Variable staggering and vertical coordinates}

>>>>

\begin{abstract}
Atmospheric dynamical cores are a fundamental component of global atmospheric modeling systems, and are responsible for capturing the dynamical behavior of the Earth's atmosphere via numerical integration of the Navi\'er-Stokes equations.  These systems have existed in one form or another for over half of a century, with the earliest discretizations having now evolved into a complex ecosystem of algorithms and computational strategies.  In essence, no two dynamical cores are alike, and their individual successes suggest that no perfect model exists.  To better understand modern dynamical cores, this paper aims to provide a comprehensive review of eleven dynamical cores, drawn from modeling centers and groups that participated in the 2016 Dynamical Core Model Intercomparison Project (DCMIP) workshop and summer school.  This review includes choice of model grid, variable placement, vertical coordinate choices, prognostic equations, temporal discretization, and the diffusion, stabilization, filters and fixers employed by each system.
\end{abstract}

@@

The Dynamical Core Model Intercomparison Project (DCMIP) is an ongoing effort targeting the intercomparison of a fundamental component of global atmospheric modeling systems: the dynamical core.  Although this component's role is simply to solve the equations of fluid motion (the Navi\'er-Stokes equations) governing atmospheric dynamics, there are numerous confounding factors and compromises that arise from making global simulations computationally feasible.  These factors include the choice of model grid, variable placement, prognostic equations, vertical coordinates, representation of topography, numerical method, temporal discretization, physics/dynamics coupling frequency, and the manner in which artificial diffusion, stabilization, filters and/or energy/mass fixers are applied.

To advance the intercomparison project and provide a unique educational opportunity for students, DCMIP hosted a multidisciplinary two-week summer school and model intercomparison project, held at the National Center for Atmospheric Research (NCAR) in June 2016, that invited graduate students, postdocs, atmospheric modelers, expert lecturers and computer specialists to create a stimulating, unique and hands-on driven learning environment. The 2016 workshop and summer school followed from earlier DCMIP and dynamical core workshops (held in 2012 and 2008, respectively), and other model intercomparison efforts.  Its goals were to provide a forum for discussing outstanding issues in global atmospheric models, incorporate international participation, and provide a unique training experience for the future generation of climate scientists. Special attention was paid to the role of simplified physical parameterizations, physics-dynamics coupling, non-hydrostatic atmospheric modeling, and variable-resolution global modeling. The summer school and model intercomparison project promoted active learning, innovation, discovery, mentorship and the integration of science and education.  Modeling groups were then invited to contribute model descriptions and results to the intercomparison effort for publication.

The summer school directly benefited its participants by providing a unique educational experience and an opportunity to interact with modeling teams from around the world.  The workshop is expected to have further repercussions on the development of operational atmospheric modeling systems, by giving modeling groups an opportunity to assess and intercompare their models with other advanced modeling systems.  Past and present intercomparison efforts have been leveraged by modeling groups to improve their own models, in turn leading to a positive impact on the quality of weather and climate simulations.  The workshop component of DCMIP has also advanced our knowledge of (1) the relative behaviors exhibited by atmospheric dynamical cores, (2) differences that arise among mechanisms for coupling the physical parameterizations and dynamical core, and (3) the impacts of variable-resolution refinement regions and transition zones in global atmospheric simulations.  Notably, the use of idealized test cases to isolate specific phenomena gave us a unique opportunity to assess specific differences that arise due to the choice of dynamical core.  Another important outcome of the workshop was the development of a standard test case suite and benchmark set of simulations that can be used for assessment of any future dynamical core.  The test cases introduced in the 2016 workshop build on the previous DCMIP test case suites \citep{jablonowski2008idealized, ullrich2012dynamical} with tests that now incorporate simplified moist physics.

This paper is the first in a series of papers documenting the results of this workshop.  It is intended to provide a review of the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling.  For easy reference, a list of mathematical symbols that are employed in this paper (and subsequent DCMIP papers) is given in Table \ref{tab:symbols}.  Section \ref{sec:DynamicalCores} then provides a brief overview of each of the participating models, along with a tabulation of relevant details about the dynamical core design.  The body of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  Section \ref{sec:ModelGrids} describes model grids that have been employed by modern modeling systems; section \ref{sec:VariablePlacement} describes aspects of variable placement and prognosis; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; and section \ref{sec:TemporalDiscretizations} describes temporal discretizations.  The summary and conclusions then follow in section \ref{sec:Conclusions}.  Finally, appendix \ref{sec:EquationSets} provides a comprehensive overview of the various forms the Navi\'er-Stokes equations take in dynamical cores, and has been included as a resource for dynamical core developers.

\begin{table}[p]
\caption{A standard list of symbols used throughout this paper and in the DCMIP.} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$s$ & Vertical model coordinate \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind \\
$v$ & Meridional wind \\
$w$ & Vertical velocity \\
$\vb{u}$ & 3D wind vector \\
$\vb{u}_h$ & Horizontal wind vector \\
$\vb{v}_h$ & Horizontal wind vector with covariant components \\
$\omega$ & Vertical pressure velocity  \\
$D$ & Divergence of the horizontal wind vector\\
$\zeta$ & Vertical component of relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$e$ & Internal energy \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$\rho_s$ & Pseudo-density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\theta$ & Potential temperature \\
$\theta_v$ & Virtual potential temperature \\
$\theta_{il}$ & Ice-liquid potential temperature \\
%$P_{ls}$ & Large-scale precipitation rate \\
$q$ & Water vapor mixing ratio \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
$q_i$ & General tracer mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

This section provide a brief overview of key discretization choices, along with unique features or design specifications from participating dynamical cores.  Further details on these choices can be found in subsequent sections.  Simulation results and model descriptions have been submitted from ten dynamical cores (see Table \ref{tab:Models}).  Although all models solve the Euler equations (or a close approximation of these equations), essentially every model employs a different formulation with a distinct set of prognostic variables.  The derivation of many of these formulations can be found in Appendix \ref{sec:EquationSets}.  The prognostic variables employed and horizontal discretizations for these dynamical cores are summarized in Table \ref{tab:ModelsHorizontal}.  The vertical staggering of variables and vertical coordinate choice is summarized in Table \ref{tab:ModelsVertical}.  Standard options for diffusion, stabilization, filters, or fixers along with the temporal discretization for these models is summarized in Table \ref{tab:ModelsTemporalDiffusion}.  The participant models are briefly described through the remainder of this section.

\begin{table}[p]
\caption{Participating modeling centers and associated dynamical cores that have submitted simulation results.} \label{tab:Models}
\begin{center}
\begin{tabular}{cll}
\hline Acronym & Model Name & Modeling Center or Group \\ \hline 
CSU & Colorado State University Model & Colorado State University, USA \\
DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
FV$^3$ & GFDL Finite-Volume Cubed-Sphere Dynamical Core & Geophysical Fluid Dynamics Laboratory, USA \\
FVM & Finite Volume Module of the Integrated Forecasting System & European Centre for Medium-Range Weather Forecasts \\
GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
HOMME & High-Order Method Modeling Environment & Sandia National Laboratories / University of Colorado, Boulder \\
ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie, Germany \\
MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research, USA \\
NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
OLAM & Ocean Land Atmosphere Model & Duke University / University of Miami, USA \\
Tempest & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis, USA \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.  Equation set indicates whether a model is hydrostatic (H) or non-hydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).  Only two numerical methods are represented among participating models, namely finite-volume (FV) and spectral-element (SE).  More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Equation set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
& & & & method & staggering \\ \hline
CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
DYNAMICO & H/NH {\color{red}(D)?} & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
FVM & NH (D) & $\rho_d$, $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d \theta$, ($\Phi$), $\rho_d q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
GEM & NH & $\vb{u}_h$, $w$, {\color{red}$\log \partial s / \partial \zeta$}, $\log T_v$, $q$ & Yin-yang (\S \ref{sec:grid_yinyang}) & FV & C-grid  \\
HOMME & H & $\vb{u}_h$, $\rho_s$, $T_v$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
ICON & NH & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
OLAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\theta_{il}$, {\color{red} $q$} & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Vertical staggering (detailed in section \ref{sec:VerticalStaggering}) and vertical coordinates (detailed in section \ref{sec:VerticalCoordinates}) for participating dynamical cores.} \label{tab:ModelsVertical}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Vertical Staggering & Vertical Coordinate \\ \hline 
CSU & Lorenz & fixed height \\
DYNAMICO & Lorenz & floating mass (\S \ref{sec:FloatingLagrangian}) \\
FV$^3$ & Lorenz & floating mass (\S \ref{sec:FloatingLagrangian}) \\
FVM & Co-located & (floating) height \\
GEM & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering}) & log pressure (\S \ref{sec:GEM_zeta})  \\
HOMME & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
ICON & Lorenz & fixed height \\
MPAS & Lorenz & fixed height \\
NICAM & Lorenz & fixed height \\
OLAM & Lorenz & fixed height with cut-cells (\S \ref{sec:CutCells}) \\
Tempest & Lorenz & fixed height \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Standard options for diffusion, stabilization, filters, or fixers in participating dynamical cores (detailed in section \ref{sec:DiffusionStabilization}) and temporal discretization (detailed in section \ref{sec:TemporalDiscretizations}).} \label{tab:ModelsTemporalDiffusion}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Standard options for diffusion, & Temporal discretization \\
& stabilization, filters, or fixers & \\ \hline 
CSU & 4th-order horizontal hyperviscosity & 3rd-order Adams-Bashforth (AB3) \\
DYNAMICO & {\color{red}?} & ARK232 \citep{giraldo2013implicit} \\
FV$^3$ & Divergence damping, monotonic limiting, hyperviscosity & Forward-backwards \citep{LR1997QJR} \\
FVM & Momentum dissipation, scalar diffusion, absorbing layer & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
HOMME & 4th-order horizontal hyperviscosity & KGU53 \citep{guerra2016high} \\
ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
MPAS & Smagorinsky, hyperdiffusion & Split-explicit \citep{klemp2007conservative} \\
NICAM & 3D divergence damping, Smagornisky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
OLAM & Divergence/vorticity damping & 2nd-order Adams-Bashforth, Lax-Wendroff (for tracers) \\
Tempest & 4th-order horizontal hyperviscosity & ARS232 \citep{ascher1997implicit} \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

{\color{red}[More references?]}

@@

The GEM model \citep{Girard2014} is used for operational forecasting at Environment and Climate Change Canada.  The horizontal discretization uses the Yin-Yang grid \citep{kageama2004yinyang} with Arakawa C-grid staggering.
The vertical coordinate is a unique hybrid terrain-following coordinate of a log-hydrostatic-pressure type
and the vertical discretization is based on the Charney-Phillips grid (see section \ref{sec:VerticalStaggering}).
A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.
It gives rise to an {\it{iterative}} process where each step requires the solution of a
linear system of equations that is reduced to a Helmholtz problem for one composite variable.
For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method \citep{Qaddouri2008schwarz}.
The dynamics and physics are time split.
To eliminate numerical noise, an explicit hyperviscosity is employed for wind components and tracers
via applications of the Laplacian operator.
Viscosity operations are applied after the completion of the physics time step.

@@

{\color{red}[More references?]}

The ICON model \citep{zangl2014icon} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain-following height-based Lorenz grid.  Prognostic horizontal velocities are stored as normal wind components $v_n$ at the edge mid-points of full levels.  Prognostic vertical wind $w$ is stored at the circumcenters of the triangles on half levels. The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves. Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics. One complete time step typically includes 5 dynamical sub-steps. The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport. For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells. Further divergence damping is applied to the normal wind at every sub-step. Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves. The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinsky scheme with a background 4th order Laplacian diffusion operator. For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion. The numerical methods have been chosen for high numerical efficiency, and they rely on next-neighbor communication only, thus allowing massive parallelization.

@@

\subsection{Ocean-Land-Atmosphere Model (OLAM)}

OLAM \citep{walko2008ocean, walko2008ocean2, walko2011direct} is a deep-atmosphere, fully-compressible model that solves the equations of motion in finite-volume momentum conservation form. Acoustic modes are solved explicitly in the horizontal and implicitly in the vertical by means of time splitting. Equations are discretized with a C-grid formulation on a hexagonal Voronoi mesh with optional local mesh refinement (which introduces some pentagons and heptagons). Height is the vertical coordinate, and a Lorenz vertical grid stagger is used. A unique feature of OLAM is that grid levels are horizontal and intersect topography. This avoids a number of well-documented errors associated with terrain-following grids and also eliminates the need for evaluation of coordinate transformation terms. Topography is represented as a smooth (non-stepped) surface by means of cut cells whose surfaces and volume are reduced according to the portion of each cell that is below ground. The OLAM cut-cell formulation conserves mass and momentum.

@@

OLAM optimizes by applying the spring dynamics method of \cite{tomita2001shallow} to the dual triangular mesh prior to its mapping to the Voronoi grid. When local mesh refinement is applied, which OLAM achieves in a series of one or more resolution-doubling steps, each spanning a transition zone that is three grid rows wide (Figure \ref{fig:OLAMRefinement}), the equilibrium spring length is scaled to the target grid cell size in each refinement level and is varied incrementally across the transition zone. Spring dynamics is further modified by forcing angles on the dual triangular mesh in the transition zone in order to move the triangle edges closer to the centers of the hexagon edges they intersect.

\begin{figure}
\begin{center}
\includegraphics[width=2.5in]{OLAM_Grid_Refinement.png}
\end{center}
\caption{Detail of one step of local mesh refinement used by OLAM Voronoi mesh. The transition zone is constructed by explicit topological reconnection of the grid lines, which produces pairings of heptagons (red dots) and pentagons (blue dots) along the refinement perimeter.} \label{fig:OLAMRefinement}
\end{figure}

@@

\section{Variable placement and prognosis} \label{sec:VariablePlacement}

Although the horizontal grid determines the approximate density and distribution of grid points, the placement of prognostic (and diagnostic) variables within each grid volume is still free to be specified.  Hence, in this section we turn our attention to the placement of prognostic variables on the grid, the choice of vertical coordinate, and the choice of corresponding prognostic equations.
